---
title: "viz and eda"
author: "Tianqi Li"
date: "2024-09-30"
output: github_document
---

```{r setup}
library(tidyverse)
library(ggridges)
library(patchwork)
library(readxl)
```

## Load the weather data
rnoaa allows you to access to dataset online that is aggregated
```{r}
weather_df <- 
  read_csv("weather_df.csv", na = c("NA", "", ".")) |>   # Read the unzipped CSV file
  janitor::clean_names()                      
```

# Scatter plots
Create the first scatterplots ever.
```{r}
ggplot(weather_df, aes(x = tmin, y = tmax)) + 
  geom_point()
```

Now approach, same plot
```{r}
weather_df |>
  ggplot(aes(x = tmin, y = tmax)) +
  geom_point()
```

Save and edit a plot object
```{r}
weather_plot = 
  weather_df |>
  ggplot(aes(x = tmin, y = tmax))

weather_plot + geom_point()
```

## Advanced scatterplot
```{r}
weather_df |>
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = .3, size = .8) +
  geom_smooth(se = FALSE)
```

Where you define aesthetics can matter
```{r}
weather_df |>
  ggplot(aes(x = tmin, y = tmax)) +
  geom_point(aes(color = name), alpha = .3, size = .8) +
  geom_smooth(se = FALSE)
```

Use faceting real quick
```{r}
weather_df |>
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = .3) +
  geom_smooth(se = FALSE) +
  facet_grid(. ~ name)
```

Lets make a somewhat more interesting scatterplot
```{r}
weather_df |>
  ggplot(aes(x = date, y = tmax, color = name, size = prcp)) +
  geom_point(alpha = .3) +
  geom_smooth(se = FALSE) +
  facet_grid(. ~ name)
```

```{r}
weather_df |>
  filter(name == "CentralPark_NY") |>
  mutate(
    tmax_fahr = tmax *(9/5) + 32,
    tmin_fahr = tmin * (9/5) +32,
  ) |>
  ggplot(aes(x = tmin_fahr, y = tmax_fahr)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

## Small things
```{r}
weather_df |>
  ggplot(aes(x = tmin, y = tmax)) +
  geom_smooth(se = FALSE)
```

```{r}
weather_df |>
  ggplot(aes(x = tmin, y = tmax)) +
  geom_hex()
```

```{r}
weather_df |>
  ggplot(aes(x = tmin, y = tmax)) +
  geom_point(color = "blue")
```

## Univriate plots

```{r}
weather_df |>
  ggplot(aes(x = tmin)) +
  geom_histogram()
```

```{r}
weather_df |>
  ggplot(aes(x = tmin, color = name)) +
  geom_histogram()
```
```{r}
weather_df |>
  ggplot(aes(x = tmin, fill = name)) +
  geom_histogram()
```

```{r}
weather_df |>
  ggplot(aes(x = tmin, fill = name)) +
  geom_histogram(position = "dodge")
```

how would I fix this? maybe facet?

```{r}
weather_df |>
  ggplot(aes(x = tmin, fill = name)) +
  geom_histogram() + 
  facet_grid(. ~ name)
```

maybe a density plot?
```{r}
weather_df |>
  ggplot(aes(x = tmin, fill = name)) +
  geom_density(alpha = .3)
```

```{r}
weather_df |>
  ggplot(aes(x = name, y = tmin, fill = name)) +
  geom_boxplot()
```

violin plots
```{r}
weather_df |>
  ggplot(aes(x = name, y = tmin, fill = name)) +
  geom_violin()
```

ridge plot
```{r}
weather_df |>
  ggplot(aes(x = tmin, y = name)) +
  geom_density_ridges()
```

LA univariate plots
```{r}
weather_df |>
  ggplot(aes(x = prcp, fill = name)) +
  geom_density(alpha = .3)

weather_df |>
  ggplot(aes(x = name, y = prcp)) +
  geom_boxplot()

weather_df |>
  filter(prcp > 10, prcp < 1000) |>
  ggplot(aes(x = prcp, fill = name)) +
  geom_density(alpha = .3)
```

## Saving and embedding plots
saving plots
```{r}
ggp_weather = 
  weather_df |>
  ggplot(aes(x = date, y = tmax, color = name)) +
  geom_point()

ggsave("plots/ggp_weather.pdf", ggp_weather, width = 8, height = 6)
```

embedding plots
```{r fig.width=6, fig.asp = .6}
weather_df |>
  ggplot(aes(x = tmin, y = tmax)) +
  geom_point(aes(color = name), alpha = .5)
```

```{r}
weather_df |>
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = .3) + 
  labs(
    title = "temperature scatterplot",
    x = "Minimum Temp (C)",
    y = "Maximum Temp (C)",
    color = "location",
    caption = "Weather data taken from rnoaa package for these three stations"
  ) +
  scale_x_continuous(
    breaks = c(-15, 0, 20),
    labels = c("-15C","0","20")
  ) +
  scale_y_continuous(
    limits = c(0,30),
    transform = "sqrt",
    position = "right"
  )
```

```{r}
ggp_scatterplot = 
weather_df |>
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = .3) + 
  labs(
    title = "temperature scatterplot",
    x = "Minimum Temp (C)",
    y = "Maximum Temp (C)",
    color = "location",
    caption = "Weather data taken from rnoaa package for these three stations"
  ) + 
  viridis::scale_color_viridis(name = "Location", discrete = TRUE)
```

```{r}
ggp_scatterplot +
  theme_bw() +
  theme(legend.position = "bottom")
```

```{r}
ggp_scatterplot +
  theme(legend.position = "bottom") + 
  theme_minimal()
```

```{r}
ggp_scatterplot +
  theme(legend.position = "bottom") + 
  theme_classic()
```

```{r}
ggp_scatterplot +
  theme(legend.position = "bottom") +
  ggthemes::theme_excel()
```

```{r}
weather_df |>
  ggplot(aes(x = date, y = tmax, color = name, size = prcp)) +
  geom_point(alpha = .3) +
  geom_smooth(se = FALSE) +
  viridis::scale_color_viridis(discrete = TRUE) +
  labs(
    x = "Date",
    y = "Maximum Temp (C)",
    title = "Seasonal variation in Max Temps"
  ) + 
  theme_minimal() +
  theme(legend.position = "bottom")
```

Use different datasets in different 'geom's

```{r}
central_park_df = 
  weather_df |> 
  filter(name == "CentralPark_NY")

molokai_df = 
  weather_df |> 
  filter(name == "Molokai_HI")

ggplot(data = molokai_df, aes(x = date, y = tmax, color = name)) + 
  geom_point() + 
  geom_line(data = central_park_df) 
```

## Multiple panels
```{r}
weather_df |>
  ggplot(aes(x = tmin, fill = name)) +
  geom_density() + 
  facet_grid(. ~ name)
```

```{r}
ggp_tmax_tmin =
  weather_df |>
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = .3)
  
ggp_tmax_density =
  weather_df |>
  ggplot(aes(x = tmax, fill = name)) +
  geom_density(alpha = .3)

ggp_tmax_date =
  weather_df |>
  ggplot(aes(x = date, y = tmax, color = name)) +
  geom_point() + 
  geom_smooth(se = FALSE)

(ggp_tmax_tmin+ggp_tmax_density)/ggp_tmax_date
```

## Data manipulation

```{r}
weather_df |>
  mutate(name = fct_relevel(name, c("Molokai_HI", "CentralPark_NY", "Waterhole_WA"))) |>
  ggplot(aes(x = name, y = tmax, fill = name)) +
  geom_violin(alpha = .5)
```

```{r}
pulse_df = 
  haven::read_sas("data/public_pulse_data.sas7bdat") |>
  janitor::clean_names() |>
  pivot_longer(
    cols = bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    names_prefix = "bdi_score_",
    values_to = "bdi_score")

pulse_df |>
  ggplot(aes(x = visit, y = bdi_score)) +
  geom_boxplot()
```

```{r}
pup_data = 
  read_csv("./data/FAS_pups.csv", na = c("NA", ".", "")) |>
  janitor::clean_names() |>
  mutate(
    sex = 
      case_match(
        sex, 
        1 ~ "male", 
        2 ~ "female"))

litter_data = 
  read_csv("./data/FAS_litters.csv", na = c("NA", ".", "")) |>
  janitor::clean_names() |>
  separate(group, into = c("dose", "day_of_tx"), sep = 3)

fas_data = left_join(pup_data, litter_data, by = "litter_number") 

fas_data |> 
  select(sex, dose, day_of_tx, pd_ears:pd_walk) |> 
  pivot_longer(
    pd_ears:pd_walk,
    names_to = "outcome", 
    values_to = "pn_day") |> 
  drop_na() |> 
  mutate(outcome = forcats::fct_reorder(outcome, pn_day, median)) |> 
  ggplot(aes(x = dose, y = pn_day)) + 
  geom_violin() + 
  facet_grid(day_of_tx ~ outcome)
```

